---
title: "60 Days with the ZSA Moonlander: Review"
description: |
  I spent 60 days with the ZSA Moonlander Keyboard and ultimately returned it. I wanted to like it, but couldn't make the ergonomics work for me. 
author:
  - name: Miles McBain
    url: https://milesmcbain.xyz
date: 06-20-2021
output:
  distill::distill_article:
    self_contained: false
categories:
  - keyboards
  - ergonomics
draft: true 
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Even if you're not a massive input nerd, there's a pretty good chance you've heard about the [ZSA Moonlander Ergonomic Keyboard](). It's billed as the next generation of the much vaunted [ErgodoxEZ]() lineage, and like it's predecessor it carries an eye-watering price tag: USD $365.

This post is my review of this apex input device. I spent 60 days with it before giving up on it due to steadily worsening strain injuries in my hands. Despite the ergonomics being something of a critical failure for me, I did find a lot to like about it, especially on the configuration and customisation side of things. It came frustratingly close to a device I could really get excited about. 

The rest of this review is going to examine the these aspects of the board:

1. Quality
2. Portability
2. Tweakability
3. Ergonomics
4. Support Experience

From my perspective these represent the main selling points.

The Moonlander I ordered was the black version with Kailh Box Red (linear) switches:

![]()

Some context about me: I spend most of my job writing code to do data science
work in R, during which I use a keyboard-driven workflow. I work standing 100%
of the time, and I avoid using a mouse wherever possible. For text editing I
favour either [Spacemacs](https://www.spacemacs.org/) or
[VSpaceCode](https://vspacecode.github.io/). My daily-driver keyboard for the
last 4 or so years has been the Microsoft Sculpt.

# Quality

For the most part the Moonlander felt like a premium device. Its surfaces felt
good under my hands. The key caps are subtly textured and ever so slightly
grippy. They felt weighty and durable. They typing sounds were thick and
substantial, without being excessively loud [^although a lot of this is due to my initial choice of a linear switch]. It gave a very tidy first impression.

I do have some nitpicks though:

The mechanism in the tenting legs felt flimsy. When rotating them I could feel a
gear clicking though a discrete set of positions, but the sound and feel of this
gave the impression that the gear was plastic. Unsettling metallic groans were
emitted from the springs on the allen screws when tightening them that made me
feel caution was required.


# Portability

This is an important one for me. I can't afford two Moonlanders, so I needed it
to be able accompany me in a bicycle pannier bag on my daily commute. Also
setting it up at the destination can't be a chore.

A big strike against the Moonlander in this regard is the lack of rigid case.
For a device in this price range it was a let down. What was provided wasn't
really a case so much as a thin neoprene slip that keeps the two halves, allen
key, and cables bundled together. It didn't seem like it would offer any
protection whatsoever from bumps during transit. By contrast my headphones,
which cost in the same ballpark as the Moonlander, shipped with a rigid zip-up
case.

After some careful measuring I was able to find exactly one reasonably priced rigid case for DJ headphones that would accept it, for which I paid AUD $40.

In its stock neoprene travel bundle, the board, cables, and allen key weighed in
at around a kilo. I didn't notice the weight on the bike, but you would in a
backpack, particularly since it is dense weight. Heavy enough to notice, not
heavy enough to be an issue I think.

The setup and tear down process does feel a bit cumbersome at first. The fit of
the allen key in the tenting screws was quite tight, so I found myself always
having to concentrate to slot it in. Perhaps the most difficult thing to get
down was that the left thumb cluster screw is reverse threaded. "Lift-to-loosen"
I found myself muttering under my breath. It's the kind of thing you forget in
hurry, which I frequently am when trying to depart.

When packing up, the cables don't have a dedicated place to live in the neoprene slip which meant pack up always felt a bit haphazard. A couple of times I accidentally popped off a keycap trying to squeeze the cables into the slip along side one of the halves.

On the plus side, the wrist wrest folding mechanism is quite slick with a magnetic clasp that keeps them in travel mode. The wrist wrest hinge also gave a robust impression.

After I got the rigid case packing was greatly simplified. I started using the slip as padding only, and the case was sufficiently large that I could keep the tenting legs in my favoured position while stowing it.

So while I agree the Moonlander is portable, I don't think it's appreciably more
portable than a small keyboard (e.g. 60% form factor). Compared with other
ergonomic boards, it probably would be at the portable end, definitely more
portable than traditional offerings from Microsoft, Kinesis, et. al.

# Tweakability

## Software 

This is where the Moonlander shone for me. The software side of things is incredibly well executed. There are two main ways you can customise the Moonlander's functionality: 

* You can use [online GUI editor called 'Oryx'](https://configure.zsa.io/moonlander/layouts/default/latest/0) which seems to be backed by `git`.
  Changes are incrementally tracked and you can fork a configuration at any
  point. Oryx is focused on the key map, but it does expose a subset of deeper
  stuff that is possible in the firmware.
* You can edit the Moonlander firmware directly in C code. It uses a framework
  called 'QMK'. ZSA provide instructions on how to get going with this, and QMK
  itself has some of the nicest documentation of any open source project that I
  have ever seen.

I didn't do anything too radical but the possibilities are expansive. To give a taste for what's possible:

I swapped out my shift keys for layer toggles. This meant I could directly control the 'shifted' version of each key. With this concept I:

* Made underscore, `_`, the shift of space.
* Swapped the locations of angle brackets `<`,`>` and parentheses `(`,`)`
* Made Lshift + Rshift `{`, with `}` on hold
* Made Rshift + Lshift `[`, with `]` on hold

The idea being to move symbols that I use regularly while working to more
convenient locations - and the results felt pretty damn good! This experience
has cemented for me that deep programmability is an important facet of truly
ergonomic hardware. All of us have different preferences and physical
limitations, too many and varied to design for in hardware!

One bonus in the Moonlander's favour is that QMK is an open source firmware that
is used by a variety of exotic keyboards including kit-builds. So experience
with QMK is likely to be portable to other boards, rather than being one
manufacturer's way of doing things.

I didn't get deep into QMK but I did work with it to set faster macros for
commonly used symbol combinations like `%>%` and `<-`, as well as playing with
the acceleration parameters of mouse control mode. This was all much easier than
I expected, thanks to the excellent docs. I found a Docker config for the QMK build environment that made building my changes a breeze.

One nitpick I'd make about tweaking that the iteration loop is quite slow for
continuous parameters. These underpin some important features, for example: How
long does a key need to be held down to register as a hold? How quick does a
double tap need to be? How fast should the mouse accelerate in mouse control
mode? The only way to set these is according to personal taste, but to change a
value means recompiling the firmware and re-flashing the board. It would be cool
if the board could provide some kind of shell that would accept commands
to alter these parameters live and feel the effects ^[This nitpick is minor since, I think within QMK it would definitely be possible to assign keys to increment and decrement these parameters live - it just would be nice to have a workflow for this out of the box.].

Nitpicks aside, the Moonlander was a software tweaker's dream. I definitely want my next board to be running QMK.

## Hardware

On the hardware side there are a few avenues for tweaking. The thumb clusters
seem to have been designed to be swappable, although no other alternatives exist
at this time. I could easily imagine ZSA is planning a cluster with a pointing
device. There's also some mysterious sets of screw holes in the base of the
board, maybe for extreme tenting mode? Again there's nothing out at the moment. 

So the hardware tweaking angle I had open to me was the key switches. They're
'hot-swappable', which to my surprise meant I could literally swap them out
while the board was hot (powered). The swapping process was repetitive but not
otherwise challenging. There was a switch puller tool included for this
purpose. It felt kind of like what imagine medieval dentistry was.

I tried 3 sets of switches all up, including the Kailh Box Red switches I
ordered the board with. The Box Reds were definitely a mistake for me. I found
the travel too deep, like typing on a pillow. This lead to a lot of
fumbling with missed presses or double-presses as I struggled to get a feel for
the actuation point. I tried pushing the switches all the way down to the bottom-out point, but that felt fatiguing.

I had a much better time with some Kailh Speed Copper switches which have lower
travel and a higher actuation point. But even these felt like work compared to
what I was used to on the Microsoft Sculpt.

The hot-swap design worked really well for me since I didn't really know what I
wanted in a switch. I ordered a sampler batch and had a fun time trying them all
in different positions to get a feel for what I liked. The Moonlander taught me a lot about the influence of switches, and the hot-swap ability is a convenient feature that I would regard highly when choosing my next keyboard, although it may not be essential, having some concept of what I like now.


# Ergonomics

ZSA's marketing bills the Moonlanader as a "Next-generation ergonomic keyboard".
Apart from the customisable key layouts, the main ergonomic features of the
Moonlander that it stakes this claim on seem to be:

* Split halves for typing at more comfortable width and angle
* Staggered 'Columnar' or 'Ortholinear' key arrangement which changes home position reach angles reach distances
* Wrist rests (removable)
* Tiltable thumb cluster
* Ability to tent the board using a tenting leg and thumb cluster.

Some of these things worked for me, others did not. I expect this is a highly personal thing. 

## Split Halves

A thing that worked was typing with a much wider posture encouraged by having
the halves set around shoulder width. I have a sense that standing to type with
my wrists close together exacerbates a tendency for me to round my shoulders
forward - since it concentrates weight in front of me. With the Moonlander I
felt like I could more easily maintain a healthier posture.

## Staggered Columns

The staggered columnar arrangement presented a bit of a stumbling block
initially. It took me 7 days to get back to my previous typing speeds ^[Perhaps I had some advantage coming from a board that was already split]. I
have mixed feelings about the arrangement. I can get on board with the idea it
makes hand movements symmetrical, but I didn't find universally better as the hype would have you believe. 

The columnar arrangement makes different tradeoffs. Some keys are easier to
reach while some get harder. For example: The middle columns are offset slightly
downward from the index finger columns. Why is this is case when these columns
use the same finger? The effect of this is that (in QWERTY) `Y` and `T` become
easier to reach while `N`, a frequent key in programming, becomes a much further reach than on a staggered board.

I did notice some fatigue and pain in my right index finger and I believe the
`N` position contributed. This isn't an insurmountable problem, it could be
fixed by swapping `N` for a less common key - but [as was noted by the creator of the Workman keyboard layout](https://workmanlayout.org/), swapping keys without consideration for the interaction effects can produce poor results. Probably a better option is swapping to a new layout that has been holistically designed.

Following that line of thinking I experimented with learning DVORAK, where I
noticed another issue introduced by the staggered columns: In DVORAK the outside
keys above home are `'` and `L`. It seems that the `L` position upsets a lot of
programmers and terminal users, in part due to proximity to `S` which is
immediately below it. I.e. `ls -l` uses the pinky for all the characters other
than space in DVORAK. So what is recommended is that the ring finger is used
when reaching upward for `L` and `-`. This is unconventional but perfectly
viable on a staggered board, but felt super awkward to me with the Moonlander's
columns.

What these examples made clear to me is that layouts designed for staggered key
arrangements do not necessarily perform as well with the offset columnar
arrangement. Which seems kind of obvious now, but was not something I
considered when purchasing the Moonlander. `N` was a significant niggle for me
in QWERTY as I suspect it may be for other programmers.

## Wrist rests

Moving along to other ergonomic features brings us to the wrist rests. They're
premium feeling plastic, but I found them too hard. I couldn't use them for much
more than about 45 minutes at a stretch before my wrists and palms started going
numb and tingly. There's a chance I might have been using them wrong, but I am
able to use the slightly softer wrist rest on my Microsoft Sculpt all day
without consciously thinking about it. I removed the wrist rests and went with
hover typing after the first few days. Thankfully this was easy to do.

## Tenting and Thumb Cluster

So now finally we arrive at the cluster that was tenting and the thumb cluster. This is a challenging part to write since I still have a lot of really raw frustration and anger attached to my experience with this part of the keyboard. 

I'll start by pointing you to how the thumb cluster is marketed by ZSA:

> The thumb module tilts up and down. Use it to find the perfect typing angle, and the wrist support module will automatically follow.

> For small hands, tilt the thumb module up. For bigger hands, tilt it down.

> A custom fit for your hand size.

Next I should tell you my hand size. A common way to measure this seems to be hand length and breadth. So here are these dimensions for me against the [2012 US Army Anthropometric Survey](http://mreed.umtri.umich.edu/mreed/downloads.html#ansur) for males:

```{r, fig.cap = "Compared with surveyed servicemen in the sample I am at the 60th percentile for length but only the 6th for breadth."}
library(tidyverse)
ansur_2 <- read_csv("./ANSUR II MALE Public.csv")

my_hand <-tibble(handlength = 195, handbreadth = 81, label = "Me")

ggplot(data = ansur_2, aes(x = handbreadth, y = handlength)) +
  geom_point(colour = "grey") +
  geom_density_2d() +
  geom_point(data = my_hand, colour = "red") +
  theme_minimal() +
  geom_text(data = my_hand,
            aes(label = "Me"),
            colour = "red",
            nudge_x = 0,
            nudge_y = 1,
            size = 8 
            ) +
  labs(
    title = "My hand size vs males in the ANSUR II 2012 sample",
    subtitle = "Contours depict observation density",
    x = "Hand breadth",
    y = "Hand length"
  )
```

So I guess I am outlier-ish. I certainly don't have 'large' hands, but to call
them 'small' could be misleading depending on what dimensions are important. I
have 'skinny' hands may be better ^[For someone in the US Army. Against warriors of the keyboard kind I expect I'd fare a little better].

With these hands I couldn't find a way to use the thumb cluster without pain.
