---
title: "Adding RStudio Addins to VSCode"
description: |
  A short description of the post.
author:
  - name: Miles McBain
    url: https://milesmcbain.xyz
date: 10-18-2020
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I recently completed a project to add RStudio addins to the [Visual Studio Code R extension](). This is post is about the why and the how of it.

# But Why?

Most people who follow me online know that I use [Spacemacs and ESS]() to write
R. A few things changed recently that made VSCode radically more appealing to me
at work, where I am forced to use Windows:

1. I learned of a Spacemacs inspired extension for VSCode: [VSpaceCode](). ^[Thanks to the [Jack of Someaa YouTube channel]()].
2. The whole 2020 and working from home thing has meant my team have been using VSCode's slick [Live share feature]() to do real-time remote pair programming and debugging. It's shockingly good.

So I started experimenting to see how close I could make VSpaceCode feel to my
Spacemacs setup. The first major sticking point was an R package of mine I use many times daily: [`fnamte`]()

To use `fnmate` your editor needs to interface with your R session to pass back
the context of your document around your cursor. R does the magic once it has
that context. 

Emacs and RStudio facilitate this in different ways, but ultimately they both
allow you execute code on data taken from the text editor and shove the result
of that into an R function called in your R session. ^[In Emacs that code is
Lisp. In RStudio it's R.]

VSCode was without this ability. So I thought that to implement a binding for my
beloved `fnmate` my options were:

  * Write my own VSCode extension - having to redo work already in the R
    extension, relating to interfacing with the R process in terminal.
  * Get a binding for `fnmate` merged into the R extension.

And I guess I was mulling over what unsatisfying options these were when I
realised that what the VSCode R extension really needed was something like the
RStudio API - HOLD ON. Can I port the RStudio API? Can I make `datapasta` and
other popular addins work with VSCode with no changes to their code?! I think I know how to do this!

It was like as soon as the lightning bolt struck I knew I was not going to be able to stop myself from doing this.

# Greenlights

I was pretty jazzed about the idea and it occurred to me that it was so good
that I was probably not the first person to have it. A GitHub search
revealed that my intuition was correct: In May of 2020, VSCode-R extension
maintainer [Andrew Craig]() had created an issue discussing the possibility. 

When I read through the discussion I saw that it had kind of stalled out because
the maintainers were worried about RStudio's reaction and legal issues. They'd
resolved to contact RStudio before making any moves.

The implementation I had in mind side stepped of their initial legal concerns,
and I was pretty confident that the RStudio team I knew, the great people that I
had met, would not be bothered.

I offered to contribute and was invited to start working on the idea.^[I'd made one PR for a small feature to the VSCode-R extension in the past, so I think that helped in this regard.] Once I had a proof of concept - 1 working API call - I decided to email the authors of `{rstudioapi}` to let them know what we were planning to do.

The RStudio response was all class. You can see for yourself in the [issue
thread](https://github.com/Ikuyadeu/vscode-R/issues/302). JJ Allaire wrote back
with encouraging words, a heads up on future API directions, and an offer to
talk about the API. Kevin Ushey did likewise, filling out some more detail on
future plans, and invited questions and feedback on the API. When I did have a
question about the API, Kevin answered it quickly.

This is RStudio walking the walk of being a company that works in the
community's interests. Diversity is good. Friendly competition is good. It's
really impressive stuff, and it's a part of this story that I really wanted to
share, because it meant I could really dive into the work unencumbered by worry
about how it would be received.

# The how

## The humane punching of ducks

Now we get to some fun stuff. How is it that I can call: `rstudioapi::getActiveDocumentContext()` from within an R session in VSCode and get back the context of my VSCode active text editor window?

The trick is one that is often used the testing context, there it's known as
'mocking'. In other context it can be called ['duck punching' or 'monkey
patching'](https://en.wikipedia.org/wiki/Monkey_patch):

Within your VSCode sessions only, I've replaced the definition of the
`getActiveDocumentContext` function in the `rstudioapi` name space, with code
that talks to VSCode instead of RStudio. The incredible thing about this to me,
is that even though it feels a bit dodgy, everything you could want to have to
do it is built into the R language.

The first piece is a function called `assignInNamespace` which allows us to do this:

```{r, echo = TRUE}
assignInNamespace("getActiveDocumentContext",
                  function(...) {
                    print("Duck punch!")
                  },
                  "rstudioapi"
                  )
```

Calling `rstudioapi::getActiveDocumentContext()` would now return "Duck punch!".

The second piece allows us to make that assignment happen, immediately after the `rstudioapi` namespace gets loaded into your R session, whenever, if ever, that may be. It's a function called `setHook`:

```{r}
rstudioapi_hook <- function() {
  print("running your rstudioapi_hook...")
  assignInNamespace("getActiveDocumentContext",
                    function(...) {
                      print("Duck punch!")
                    },
                    "rstudioapi"
                    )
}

setHook(packageEvent("rstudioapi", "onLoad"), rstudioapi_hook)

rstudioapi::getActiveDocumentContext()
```

This is really nice since it means if the user never uses the `rstudioapi` we
never do anything spooky. It's also handy to be able to patch the API functions
selectively, since it allows us to keep `rstudioapi` functions that don't talk
to the IDE unchanged, e.g. `rstudioapi::is.document_range`, and we can call
those in our adapted versions.

## VSCode-R communications

The VSCode-R extension already implemented a one-way communication channel
between the user's R session and VSCode for signalling custom rendering
behaviour of plots, shiny apps, web pages etc. So my main challenge was to
implement the return protocol so that VSCode could send data back to the R
session in response to `rstudioapi` requests. This had to be written in
Typescript using the VSCode API.

The way in which the communications worked was quite interesting to me - I guess
I assumed it would be some kind of fancy web socket protocol thing. Actually
it's pretty lofi, based on files. It worked like this:

1. On startup VSCode R sessions create two files `request.log` and
   `request.lock`. In VSCode/Typescript land a file watcher process is attached
   to the lock file that triggers a callback function whenever the lock file is
   updated.
2. In the R session, a request for VSCode is made by writing some JSON to the
   `request.log` and then writing the current timestamp to the `request.lock`.
3. In VSCode the callback for the lock file update fires and confirms that the
   timestamp indicates a new message, and fires the JSON off to a request
   router.
4. A request router is just a fancy name for a giant switch statement that
   examines the JSON and decides what functions to call.

My implementation for the return path copied these ideas. Once you make an
`rstudioapi` call from R, the R sessions blocks, checking a `response.lock` file
every 100ms until an update appears and response data is read from
`response.log`.

One mistake I made initially was that I thought some API calls might be able to
be asynchronous, for example inserting text into a document - it seems like it
could done without a response back to R. But I soon learned in testing that
there are addins out there that blast the `rstudioapi` with a stream of function
calls ^[Thankyou `{remedy}` and `{sinew}` ;)], and I needed to make sure the all
changes in VSCode had been applied before I allowed a new request to be written,
otherwise request data might be overwritten without being handled.


# Challenges

## Typescript workflow

Typescript is a statically typed Javascript-a-like that compiles to Javascipt. I
quite enjoy the language, but I found working without a REPL to be
excruciatingly slow initially.

I was going:
  
  * write code
  * compile and build extension
  * test extension
  * repeat
  
And the compilation/build time, and repetitive nature of setting up state for a test was a drag. A couple of things I discovered:

  * If you put a breakpoint inside a function you are developing you can get to
    something like a REPL pretty quickly by doing your development in debug
    mode. You have the entire application's state available in an object called
    `vscode_1` that you can tinker with.
  * By stopping fighting the type checker, and refactoring code that didn't play
    nicely with it, I found the strong linting and hover help capability of
    VSCode gave me enough context to work without the safety net of the REPL.
    It's a different kind of safety net, one that works well with application code.

## Merging a large PR

The maintainers of the VSCode-R extension have quite high standards for the
code. [Kun Ren]() gives unrelentingly thorough code reviews, which are actually
a really valuable thing that I haven't had a lot of in my software-writing
career. It felt a little like a session with a code personal trainer. I
realised I had been lazy in parts, or there were nicer ways of doing things that
weren't as complex as I had estimated.

I was quite sympathetic to this process. Asking people to take on 1000 lines of
potential technical debt is a big deal. We're playing catch with a loaded
foot-gun - whatever precautions the catcher wants to take have to be respected.


# And then I broke library(tidyverse)

It turned out that much to my embarrassment the foot-gun went off pretty much
immediately!

When repository owner [Yuki Ueda] creates a release, it automatically flows
through to VSCode users the next time they open VSCode. The PR was merged in the
afternoon and in the evening I logged into my work machine to catch up on a task
from a few hours earlier. My code fell in a heap at `library(tidyverse)` with an
error that was painfully familiar, since I coded the message myself:

```
error: This `{rstudioapi}` function is not currenlty implemented for VSCode.
```

WAT. I frantically tried to reproduce the bug on my personal laptop running
Ubuntu and couldn't. Ken Ren couldn't reproduce it on Mac OS either. It seemed
we had missed some Windows specific behaviour which I was the first to encounter
on my Windows work machine.

In the midst of this, a colleague of mine who is studying messaged me to let me know the bug had sent someone in his study group into a panic and he had bailed them out with a fork of the extension. I felt terrible!

Over the next couple of hours we worked out the issues and implemented a fix. We
also made the `{rstudioapi}` emulation in VSCode opt-in for now with an option
that must be set: `options(vsc.rstudioapi = TRUE)` in case there are any other
big showstoppers we haven't seen yet. 

In retrospect the option would have been a good idea from the start, I did think
about it at some point, but I guess I got overconfident.

The bug itself is was quite interesting to me. When a user does
`library(tidyverse)` for the first time in a session there is an on-attach hook
that runs that prints a familiar _fancy_ looking message:

```
library(tidyverse)
── Attaching packages ─────────────────────────────────────── tidyverse 1.3.0 ──
✔ ggplot2 3.3.2     ✔ purrr   0.3.4
✔ tibble  3.0.3     ✔ dplyr   1.0.2
✔ tidyr   1.1.2     ✔ stringr 1.4.0
✔ readr   1.3.1     ✔ forcats 0.5.0
── Conflicts ────────────────────────────────────────── tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
```

This reproduction on my blog is not as colourful as the one we see. 

This message makes heavy use of two packages for command line output: `{crayon}`
and `{cli}`. Now the first time _these_ packages are called in a session they
fire off hundreds of lines of code designed to sniff details of your terminal
and operating system, ultimately to determine:

* What is the number of colours in the palette I can use in the terminal?
* What symbols are supported in the terminal?

On Windows only, `{cli}` ends up needing to know the version of RStudio you're
running, and it gets this via `rstudioapi::getVersion`. Up until this point
I had shied away from _spoofing_ an API version - that is my implementation
would return `TRUE` for `rstudioapi::isAvailable()` but error if you tried to
get the version or assert a specific version. This worked well for the dozen or so addin packages I tested.

There was no way around VSCode pretending to have an RStudio version for now to get `tidyverse` working. We shall see if this creates any issues. 

# Conclusion
